|

### 2021-10-13 05-Chp-CPRD-Analysis.Rmd
End of day

# Primary analysis of lipid regulating agents and dementia outcomes {#cprd-analysis-heading}

@@

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/helper.R")
knitr::read_chunk("R/04-Code-CPRD-Analysis.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

@@

<!-- TODO Need to add some information on the generation and use of propensity scores -->

@@

<!-- TODO Check lipid-regulating vs lipid regulating -->

@@

Electronic health record (EHR) databases are large collections of patients data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standard set of codes. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the length of time for which participants are followed. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

This analysis makes use of the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of more than 3 million people from general practices across the UK. Using this data, the analysis presented in this chapter examined the effect of treatment which lower cholesterol levels, such as statins, on the risk of dementia and related outcomes.

Little evidence for an effect of lipid regulating agents effect on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of drugs called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia with lipid regulating agent use.

This increased risk is unexpected, and is very likely to be due to the presence of bias in the analysis. This bias, called "confounding by indication", is caused when those who are prescribed a statin are more at risk of vascular dementia for a range of reasons, and so it appears as if statins are harmful. However, despite this limitation, the analysis presented provides an important source of information which will be used in later chapters.

@@

In this Chapter, I present the results of an analysis of a large population-based electronic health record dataset to investigate the relationship between lipid regulating agent (LRA) use and dementia outcomes.

The analysis aims to address important two limitations of the current evidence base as identified by the systematic review presented in Chapter \@ref(sis-rev-heading).

Firstly, it explicitly examines vascular dementia as an outcome. The systematic review presented in the previous chapter identified an evidence gap around the effect of lipid regulating agents on the risk of vascular dementia. As triangulation exercises require as many diverse sources of evidence as possible, this analysis provides a source of information on this outcome.

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia (as identified by the systematic review). Specifically, this involved a concerted effort to address immortal time bias though use of a Cox Proportional time-varying treatment indicator. This approach provides a evidence source at risk of a distinct set of biases due to the alternative analytical strategy that will be incorporated into the triangulation exercise presented in Chapter \@ref(discussion-heading).

@@

Previously known as the General Practice Research Database (GPRD), the Clinical Practice Research Datalink (CPRD) is a large population-based, electronic health record (EHR) database.[@herrett2015] The database has been collecting primary care data from participating practices across England since 1987.[@williams2012; @wood2001revitalizing] It contains the primary care records for more than 10 million primary care patients in England, and is broadly representative of the UK population in terms of age, sex and ethnicity.[@herrett2015; @mathur2014]

@@

Lists of relevant codes for each of the index events, exposures and outcomes used in the analysis were created. Each of these are described in more detail in the following sections.

@@

This analysis included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cutoff were included in the original CPRD extract obtained for this analysis. However, these were excluded from the analysis as data quality and reliability is thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

@@

All events of interest were identified using predetermined code lists, which are available for inspection from the archived repository accompanying this analysis (data/code availability is discussed in Section \@ref(cprd-data-avail)).

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); or last registration date with their GP practice. Participants were removed from the sample if they were less than 40 years of age, had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

@@

A participant’s drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. We did however examine how often the initial drug class altered according to one of three criteria:

@@

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome. When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (see Figure \@ref(fig:decisionTreeFig)). The diagnosis date of the outcome was determined by the first record of a relevant code.

<!----------------------------------------------------------------------------->
(ref:decisionTreeFig-cap) Decision tree for assigning dementia subtypes, based on the presence of Read codes in the patient's record. Note that an outcome of "Probable" or "Possible" Alzheimer's disease (AD) requires the absence of any vascular outcome codes.

@@

Participants were censored at the earliest of date of: diagnosis; death; transfer out of the study cohort; or end of follow-up (29th February 2016).

@@

A range of additional variables were included in the analysis, intended to address the different distributions of potential confounding variables between those who were prescribed an lipid-regulating agent and those who were not.

Demographic covariates adjusted for included age and gender. Age was calculated at date of entry into the cohort. Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived.  Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

@@

In this analysis, missing data were handled using a multiple imputation approach. Variables with missing observations were identified, and 20 imputed datasets were created.[@sterne2009] Nominal variables with missing values were modelled using multinomial logistic regression, while continuous variables were modelled using linear regression. As per best practice, all variables used in the analytic model, including the outcome, were included in the imputation model.[@moons2006using] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) command in STATA16.

Missing data was only considered an issue for variables where a numerical test result was expected (e.g. BMI), or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition, rather than being considered missing.[@wells2013strategies]

Using multiple imputation to handle missing data is an alternative to a "complete case" approach,[@pigott2001review] where participants missing any covariate are dropped from the dataset. As a sensitivity analysis, I preformed and compared the results of both methods,[@hughes2019] to investigate the impact of multiple imputation on the results.

@@

A Cox proportional hazards (PR) model was used to estimate the effect of statins on dementia outcomes. Cox PR models are defined, in general terms, as:

@@

As the values of $b_i$ increases, value of $exp(b_i)$ (i.e. the hazard ratio) increases in tandem, thus reducing the survival time, $t$.

A Cox PR model was chosen for this analysis as it inherently accounts for the length of time participants spend in each exposure group. Using this approach, time-at-risk can be properly attributed, thus mitigating the impact of immortal time bias. This is discussed in detail in the following section.

@@

<!-- IDEA Need to describe accounting for clustering here: id it somethings I did? See other thesis for clustering by practice -->

@@

Immortal time bias covers two distinct but related types of bias. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to the exposure is excluded leading to the exposed and control groups being followed up from different time points.[@levesque2010] For example, if the unexposed group are followed from a cholesterol test result, while the exposed group is followed from date of LRA prescription, any events that occur in the exposed group prior to the exposure event will be inappropriately excluded from the analysis.

The second presentation of immortal time bias is as a type of misclassification bias (Panel B, Figure \@ref(fig:immortalTimeBias)). It occurs when the exposure time prior to the exposure date, and any events occurring within it, is inappropriately assigned to the exposed group. This second presentation appears to be common in the existing literature, as several of the studies included in the systematic review presented in Chapter \@ref(sys-rev-heading) were identified as being at risk of immortal time bias following formal risk of bias assessment using the ROBINS-I tool (see Section \@ref(risk-of-bias-subheading)).

This analysis is primarily concerned with the second presentation of immortal time bias, as all participants were followed from a common index date (earliest of: date of raised cholesterol test results; hypercholesterolemeia diagnosis; or LRA prescription). To address the potential for this second form of immortal time bias in the analysis, I employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010]

Under this approach, all patients start in the unexposed group, & contribute time-at-risk until they are prescribed a lipid regulating agent and move into the exposed group. Note, patients for whom prescription of a lipid regulating agent was the index event only contribute time to the exposed group (i.e. they enter the cohort and move into the exposed group on the same day).

<!--------------------------------------------------------------------------->
(ref:immortalTimeBias-cap) Diagram illustrating the two presentations of immortal time bias, as a selection bias (Panel A) and a misclassification bias (Panal B).

@@

```{r immortalTimeBias, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:immortalTimeBias-cap)', out.width='100%', fig.scap='(ref:immortalTimeBias-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","immortal_time.png"))
```
<!--------------------------------------------------------------------------->

### Time axis

@@

The primary analysis examined the effect of a lipid regulating agent on dementia risk, stratified by outcome and drug class. To assess the robustness of the results, a number of sensitivity analyses were performed. These are described in the following sections.

@@

In addition to the primary outcomes of interest (described in Section \@ref(cprd-outcomes)), I extracted data on two additional control outcomes. The inclusion of control outcomes in observational analyses are a useful technique to assess the strength of uncontrolled confounding,[@lipsitch2010] and these outcomes are usually class as either "negative" or "positive" outcomes. Negative outcomes are those without a likely causal path between the exposure and outcome (see Figure \@ref(fig:negativeOutcome) for a directed acyclic graph describing an ideal negative outcome).

Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are useful in observational epidemiology, as if the analysis can reproduce a known result for the control outcome, confidence in the result for the outcome of interest is increased.

Due to the wealth of data available on statins as a lipid regulating agent, control outcomes were chosen in reference to this drug class. Muscular backpain was chosen as a negative control outcome in this analysis. Despite observational analyses suggesting a link between statins and muscle pain (as opposed to more serious complications such as myopathy),[@selva-ocallaghan2018] systematic reviews of the adverse events of statin use[@collins2016a] and N-of-1 trials explicitly exploring the association of statin use with muscle pain[@herrett2021] have found little evidence supporting an effect. As such, if statins are not associated with backpain in this analysis, then confidence in the results for the dementia outcomes is increased.

Additionally, incident ischemic heart disease was included as a positive control outcome, given the well-established protective effect of lipid-lowering treatment, via statins, on the risk of this condition.[@collins2016a] Similar to the backpain outcome, if the analysis strategy can reproduce this known protective association, this will provide evidence that potential confounders have been sufficiently adjusted for.

<!-- TODO Need to add bit to this section and to results on outcome for Type 2 Diabetes outcome -->

<!----------------------------------------------------------------------------->
(ref:negativeOutcome-cap) Causal diagram showing relationship between exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of any arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders $U$ (assuming $C$ has been adjusted for).

(ref:negativeOutcome-scap) DAG of ideal negative outcome

```{r negativeOutcome, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:negativeOutcome-cap)', out.width='100%', fig.scap='(ref:negativeOutcome-scap)', message=FALSE, warning = FALSE}
library(dagitty)
library(ggdag)

ggplot2::ggsave(filename = file.path("figures","cprd-analysis","negativeOutcome.png"),
  dagitty::dagitty("dag {
    U -> N
    C -> N
    X <- U -> Y
    X <- C -> Y
    X -> Y
    bb =\"-2.368,-2.972,2.597,2.726\"
    C [pos=\"-0.3,0\"]
    N [pos=\"0.5,1\"]
    U [pos=\"-0.15,1.5\"]
    X [exposure,pos=\"0,-1\"]
    Y [outcome,pos=\"0.5,-1\"]
  }"
) %>% ggdag_classic() +
  theme_dag()

@@

```
<!----------------------------------------------------------------------------->

@@

To observe the effect of adjusting for additional covariates, I ran a model unadjusted except for age (captured via the time axis in the Cox model) and gender, and compared the results with the full adjusted model.

@@

Two sensitivity cohorts were also created. The first stratified by year of entry into the cohort in an attempt to assess for time period effects. The second removed participants who may have been pregnant  (coded as under 55 and female) to assess the robustness of the estimates, as statins are contraindicated in pregnancy, .[@karalis2016]

@@

#### Impact of dementia code lists

As part of an exploratory analysis of the effect of the choice of code lists on the analysis, I created an alternative Alzheimer's disease and non-Alzheimer's dementia outcome using code lists from a previous study published in 2009 by Smeeth _et al_.[@smeeth2009]

This previous study used a propensity matching approach to estimate the association of statins with a range of outcomes, and found evidence for a protective effect of statin use on all-cause dementia (`r estimate(0.81, 0.69,0.96,type = "HR")`) and non-AD dementia (`r estimate(0.82,0.69, 0.97,type = "HR")`), but little evidence of an effect on AD (`r estimate(0.81,0.49, 1.35, type="HR")`).

The code lists used were obtained through correspondence with the authors of that study, and are available for inspection (see Section \@ref(cprd-data-avail)).

@@

@@

<!--------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) Attrition of CPRD participants as the eligibility criteria were applied. Most attrition was due to the absence of an index event of interest.

@@

The median participant age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r comma(p1[1,2])` patients  (`r comma(p1[2,2])` probable AD, `r comma(p1[3,2])` possible AD, `r comma(p1[4,2])` vascular dementia, `r comma(p1[5,2])` other dementia: Table \@ref(tab:followUp-table)). The distribution of baseline characteristics across the seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

\blandscape
<!--------------------------------------------------------------------------->
(ref:cprdCharacteristics-caption) Patient characteristics by drug class. Summary statistics are presented as "% (N)" unless otherwise specified in the variable name.

@@

A substantial majority (`r percentage.statins`) of participants prescribed a lipid-regulating agent were prescribed a statin. I excluded the "Ezetimibe and statins" and "Nicotinic acid groups" classes from subsequent analysis based on the extremely small number of participants in these groups (N=127 & N=168 respectively; see Table \@ref(tab:cprdCharacteristics-table)). Note that the "Ezetimibe and statins" treatment group represent those prescribed a single treatment containing both ezetimibe and statins, rather than those where the two treatments were prescribed concurrently.

@@

<!-- IDEA Described the average length of time between patients being prescribed a statin and being diagnosed with dementia -->

<!--------------------------------------------------------------------------->
(ref:cprdSSA-caption) Participants who stopped, switched or added treatments by initial treatment type.

@@

```{r cprdSSA-table, message=FALSE, results="asis", echo = FALSE}
```
<!--------------------------------------------------------------------------->

@@

Crude rate ratios per 100,000 participant-years-at-risk were calculated for each outcome and class of interest and are presented in Table \@ref(tab:followUp-table).

<!----------------------------------------------------------------------------->
(ref:followUp-caption) Summary

(ref:followUp-scaption) followUp

```{r followUp-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->

\elandscape

@@

For each outcome, the overall "Any drug" estimate was driven by the statin subgroup, based on it's large size relative to the other drug classes.

@@

My results show litte evidence was found for an effect of lipid-regulating agents on probable (`r probad_text`) and possible (`r possad_text`) Alzheimer's disease when compared with no treatment, with the sole exception of fibrates on probable Alzheimer's disease (`r probad_fib_text`).

@@

In contrast to the findings for Alzheimer's disease outcomes, lipid-regulating agents were associated with an increased risk of a subsequent diagnosis of vascular dementia (`r vasdem_text`) or other dementia (`r othdem_text`). Again this effect was driven mainly by the statin subgroup, but there was some evidence that ezetimibe was associated with an increased risk of vascular (`r vasdem_eze_text`) and other (`r othdem_eze_text`) dementia.

@@

<!----------------------------------------------------------------------------->
(ref:completeCaseFig-cap) Comparison of complete case versus imputed data analyses.

>>>>

# Primary analysis of lipid-regulating agents and dementia outcomes {#cprd-analysis-heading}

@@

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/helper.R")
knitr::read_chunk("R/05-Code-CPRD-Analysis.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

@@

@@

@@

Electronic health record (EHR) databases are large collections of medical data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standardised coding system. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the length of time for which participants are followed. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

This analysis makes use of the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of more than 3 million people from general practices across England. Using this data, the analysis presented in this chapter examined whether treatments which lower cholesterol levels (also known as lipid-regulating agents or LRA) of which statins are a prime example, affect the risk of all-cause dementia and related outcomes (Alzheimer's disease, vascular dementia and other dementias).

Little evidence for an effect of lipid-regulating agents effect on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of lipi-regulatin agent called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia was associated with lipid-regulating agent use.

This increased risk in outcomes with a vascular element (e.g. vascular dementia) is unexpected, and is very likely to be due to the presence of bias in the analysis. This bias, called "confounding by indication", is caused when those who are prescribed a statin are more at risk of vascular dementia for a range of reasons, which makes it appear as if statins are harmful. Despite this limitation, the analysis presented provides an important source of information which will be used in later chapters.

@@

In this Chapter, I present an analysis of a large population-based electronic health record dataset to investigate the relationship between lipid-regulating agent (LRA) use and dementia outcomes. The analysis aims to address important two limitations of the current evidence base as identified by the systematic review presented in Chapter \@ref(sis-rev-heading).

Firstly, it explicitly examines vascular dementia as an outcome. The systematic review presented in the previous chapter identified an evidence gap around the effect of lipid-regulating agents on the risk of vascular dementia. As triangulation exercises require as many diverse sources of evidence as possible, this analysis provides a source of information on this outcome.

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia as identified by the systematic review. Specifically, this involved a concerted effort to address immortal time bias though use of a Cox proportional hazards analysis, incorporating a time-varying treatment indicator.[@suissa2008] This approach provides a evidence source at risk of a distinct bias due to the alternative analytical strategy. The results from this analysis will be incorporated into the triangulation exercise presented in Chapter \@ref(discussion-heading).

@@

Previously known as the General Practice Research Database (GPRD), the Clinical Practice Research Datalink (CPRD) is a large population-based electronic health record (EHR) database.[@herrett2015] The database has been collecting primary care data from participating practices across England since 1987.[@williams2012; @wood2001revitalizing] It contains the primary care records for more than 10 million primary care patients in England, and is broadly representative of the UK population in terms of age, sex and ethnicity.[@herrett2015; @mathur2014]

@@

The index events, exposures and outcomes used in this analysis were identified using predetermined code lists, which are available for inspection from the archived repository accompanying this analysis (data/code availability is discussed in Section \@ref(cprd-data-avail)).

<!-- TODO, ensure code list position is accurately documented in  -->

@@

This analysis included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cut-off were included in the original CPRD extract obtained for this analysis. However, these were excluded from the analysis as data quality and reliability are thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

@@

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: (a) an outcome of interest; (b) death; (c) end of follow-up (29 February 2016); (d) last registration date with their GP practice; or (e) the last CPRD collection date for their practice. Participants were ineligble for our cohort if they were less than 40 years of age (as these patients are are less likely to be prescribed a LRA), had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

@@

A participants drug class was assigned based on their first recorded prescription, and any drug switching was ignored in an effort to mimic an intention-to-treat approach. We did however examine how often the initial drug class altered according to one of three criteria:

@@

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementias, and a composite all-cause dementia outcome. When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (Figure \@ref(fig:decisionTreeFig)). The diagnosis date of the outcome was determined by the first record of a relevant code.

<!----------------------------------------------------------------------------->
(ref:decisionTreeFig-cap) Decision tree for assigning dementia subtypes, based on the presence of Read codes in the patient's record. Note that an outcome of "probable" or "Possible" Alzheimer's disease (AD) requires the absence of any vascular outcome codes.

@@

@@

A range of additional variables were included in the analysis, intended to address the different distributions of potential confounding variables between those who were prescribed an lipid-regulating agent and those who were not. These are discussed in detail below and summarised in Table \@ref(tab:covariateDef-table).

(ref:covariateDef-caption) Definition of covariates adjusted for in the Cox PR model. The codelists used to define the majority of these covariates were originally created for use in a previously published analysis.[@walker2020] while otherss were built on or adapted from previous published work,[@khan2010; @taylor2016; @wright2017].

```{r covariateDef-table, echo = FALSE}
```

Demographic covariates adjusted for included age and gender. Age was calculated at date of entry into the cohort, and adjusted for via its use as the time axis for the Cox model (see Section \@ref(cprd-time-axis)). Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived.  Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

@@

In this analysis, missing data were handled using a multiple imputation approach.[@sterne2009] Variables with missing observations were identified for inclusion in the imputation model. Nominal variables with missing values were modelled using multinomial logistic regression, while continuous variables were modelled using linear regression. As per best practice, all variables used in the analytic model, including the outcome, were included in the imputation model.[@moons2006using] Using the MICE (Multiple Imputation by Chained Equations) command in STATA16, 20 imputed datasets were created.

Missing data was only considered an issue for variables where a numerical test result was expected (e.g. BMI), or where a code existed for the absence of the condition (e.g. categorical smoking status). This approach was necessary, as absence of a code for other treatments or conditions (e.g. statin use or dementia) was assumed to indicate absence of the treatment/condition rather than being considered missing.[@wells2013strategies]

@@

A Cox proportional hazards (PR) model was used to estimate the effect of statins on dementia outcomes. Cox PR models are defined in general terms as:

@@

A Cox PR model was chosen for this analysis as it inherently accounts for the length of time participants spend in each exposure group. Using this approach, time-at-risk can be properly attributed to the appropriate exposure group, thus mitigating the impact of immortal time bias. This is discussed in detail in the following section.

@@

<!-- IDEA Need to describe accounting for clustering here: id it something I did? See other thesis for clustering by practice -->

@@

Immortal time bias describes two distinct but related types of bias. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to the exposure is excluded leading to the exposed and control groups being followed up from different time points.[@levesque2010] Following the example displayed in the figure, the unexposed group are followed from the date of a diagnosis, while the exposed group is followed from date of a prescription. In this scenario, the time between diangosis and prescription for the exposure group is missing, and any events that occur in the exposed group prior to the prescription will be inappropriately excluded from the analysis.

The second presentation of immortal time bias is as a type of misclassification bias (Panel B, Figure \@ref(fig:immortalTimeBias)). It occurs when the exposure time prior to the prescription, and any events occurring within it, is inappropriately assigned to the exposed group.

&nbsp;<!----------------------------------------------------------------------->

(ref:immortalTimeBias-cap) Diagram illustrating the two presentations of immortal time bias, as a selection bias (Panel A) and a misclassification bias (Panal B).

@@

```{r immortalTimeBias, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:immortalTimeBias-cap)', out.width='100%', fig.scap='(ref:immortalTimeBias-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","immortal_time.png"))
```

&nbsp;<!----------------------------------------------------------------------->

This second presentation appears to be common in the existing literature, as several of the studies included in the systematic review presented in Chapter \@ref(sys-rev-heading) were identified as being at risk of immortal time bias following formal risk of bias assessment using the ROBINS-I tool (see Section \@ref(risk-of-bias-subheading)). Both presentations of immortal time bias appear in the literature on the relationship of statins and dementia, as identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading)/\@ref(sys-rev-results-heading)

<!-- TODO May need to crosslink a bit better -->

This analysis attempted to address this issue by following all participants from a common index date (defined as earliest of: (a) date of raised cholesterol test results; (b) hypercholesterolemeia diagnosis; or (c) LRA prescription). Following a recommended approach to addressing the second form of immortal time bias, I employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010]

Under this approach, all patients start in the unexposed group and contribute time-at-risk until they are prescribed a lipid-regulating agent and move into the exposed group. Note, patients for whom prescription of a lipid-regulating agent was the index event only contribute time to the exposed group (i.e. they enter the cohort and move into the exposed group on the same day).

&nbsp;<!----------------------------------------------------------------------->

### Time axis {#cprd-time-axis}

@@

The primary analysis examined the effect of a lipid-regulating agent on dementia risk, stratified by outcome and drug class. To assess the robustness of the results, a number of sensitivity analyses were performed. These are described in the following sections.

#### Complete case vs imputed data

Using multiple imputation to handle missing data is an alternative to a "complete case" approach,[@pigott2001review] where participants missing any covariate are dropped from the dataset. As a recommended sensitivity analysis,[@hughes2019] I preformed and compared the results of both methods, to investigate the impact of multiple imputation on the results.

@@

In addition to the primary outcomes of interest (described in Section \@ref(cprd-outcomes)), I extracted data on three additional control outcomes. The inclusion of control outcomes in observational analyses are a useful technique to assess the strength of uncontrolled confounding,[@lipsitch2010] and these outcomes are usually class as either "negative" or "positive" outcomes.

Negative outcomes are defined as those without a likely causal path between the exposure and outcome (see Figure \@ref(fig:negativeOutcome) for a directed acyclic graph, or DAG, describing an ideal negative outcome). Conversely, positive control outcomes are those with a known causal association with the exposure of interest, ideally sourced from large well conducted randomised controlled trials. Positive control outcomes are useful in observational epidemiology, as if the analysis can reproduce a known result for the control outcome, confidence in the result for the outcome of interest is increased. Due to the wealth of data available on statins as a lipid-regulating agent, I chose three control outcomes were chosen in reference to this drug class: back pain (negative control), ischaemic heart disease (positive protective control), and Type 2 diabetes (positive harmful control).

&nbsp;<!----------------------------------------------------------------------->  
(ref:negativeOutcome-cap) Causal diagram (directed acyclic graph) showing relationship between exposure $X$, outcome $Y$, confounders (measured $C$ and unmeasured $U$) and an ideal negative outcome $N$. Note the absence of any arrow between $X$ and $N$. In this scenario, any association observed between $X$ and $N$ is due to the presence of uncontrolled confounders $U$ (assuming $C$ has been adjusted for).

(ref:negativeOutcome-scap) Causal diagram for ideal negative outcome

```{r negativeOutcome, echo = FALSE, results="asis", fig.pos = "H", fig.align='center', fig.cap='(ref:negativeOutcome-cap)', out.width='80%', fig.scap='(ref:negativeOutcome-scap)', message=FALSE, warning = FALSE}
library(dagitty)
library(ggdag)

ggplot2::ggsave(filename = file.path("figures",
                                     "cprd-analysis",
                                     "negativeOutcome.png"),
  dagitty::dagitty("dag {
    U -> N
    C -> N
    X <- U -> Y
    X <- C -> Y
    X -> Y
    bb =\"-2.368,-2.972,2.597,2.726\"
    C [pos=\"-0.3,0\"]
    N [pos=\"0.5,1\"]
    U [pos=\"-0.15,1.5\"]
    X [exposure,pos=\"0,-1\"]
    Y [outcome,pos=\"0.5,-1\"]
  }"
) %>% ggdag_classic() +
  theme_dag()

@@

```
&nbsp;<!----------------------------------------------------------------------->

Despite observational analyses suggesting a link between statins and muscular pain (as opposed to more serious complications such as myopathy),[@selva-ocallaghan2018] systematic reviews of the adverse events of statin use[@collins2016a] and N-of-1 trials explicitly exploring the association of statin use with muscle pain[@herrett2021] have found little evidence supporting an effect. This suggests that muscular backpain would be suitable for use as a negative control outcome in this analysis. Under this approach, if statin use is found to be associated with muscular backpain in this analysis, this suggests the presence of residual confounding and reduces my confidence in the results for the dementia outcomes.

Similarly, incident ischemic heart disease and Type 2 diabetes were included as a protective and harmful positive control outcome, respectively. The protective effect of lipid-lowering treatment, via statins, on the risk of ischemic heart disease is well-established, [@collins2016a] while there is growing evidence for an increased risk of Type 2 diabetes with statin use.[@collins2016a; @macedo2014; @smit2020] Similar to the negative outcome, if the analysis strategy can reproduce these known associations, this will provide evidence that potential confounders have been sufficiently adjusted for.

&nbsp;<!----------------------------------------------------------------------->

@@

To observe the effect of adjusting for additional covariates, I ran two additional models unadjusted except for: (a) age; and (b) age and gender. The results of these models was then compared the results with the fully adjusted model.

&nbsp;<!----------------------------------------------------------------------->

@@

Two sensitivity cohorts were also created. The first stratified by year of entry into the cohort in an attempt to assess for time period effects. The second removed participants who may have been pregnant (coded as under 55) to assess the robustness of the estimates, as statins are contraindicated in pregnancy.[@karalis2016]

@@

#### Comparing dementia codelists

As part of an exploratory analysis of the effect of the choice of code lists on the analysis, I created an alternative Alzheimer's disease and non-Alzheimer's dementia outcome using code lists from a published study by Smeeth _et al_.[@smeeth2009]

This previous analysis used a propensity matching approach to estimate the association of statins with a range of outcomes in The Health Improvement Network (THIN) database, a alternative source of English electronic health records which has substantial overlap with the CPRD.[@carbonari2015] The code lists used in this analysis were obtained through correspondence with the authors of that study, and are available for inspection (see Section \@ref(cprd-data-avail)).

&nbsp;<!----------------------------------------------------------------------->

@@

```{r cprdCharacteristics-setUp, echo = FALSE}
```

@@

<!--------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) Attrition of CPRD participants as the eligibility criteria were applied. The largest cause of attrition was the absence of an index event of interest.

@@

The median participant age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r comma(p1[1,2])` patients  (`r comma(p1[2,2])` probable AD, `r comma(p1[3,2])` possible AD, `r comma(p1[4,2])` vascular dementia, `r comma(p1[5,2])` other dementias).

The number of events, time-at-risk and crude rates for each drug class, tabulated by dementia outcome, are shown in Table \@ref(tab:followUp-table). A substantial majority (`r percentage.statins`) of participants prescribed a lipid-regulating agent were prescribed a statin. I excluded the "Ezetimibe and statins" (n=`r eze_sta_n`) and "Nicotinic acid groups" (n= `r nag_n``) classes from subsequent class-based subgroup analyses based on the extremely small number of participants in these groups. Note that the "Ezetimibe and statins" treatment group represent those prescribed a single treatment containing both ezetimibe and statins, rather than those where the two treatments were prescribed concurrently.

<!----------------------------------------------------------------------------->
(ref:followUp-caption) Summary of number of events, years at risk and crude rates per 100,000 participant-years-at-risk stratified by dementia outcome and drug class of interest.

(ref:followUp-scaption) Crude rates, stratified by outcome and drug class of interest.

```{r followUp-table, message=FALSE, results="asis", echo = FALSE}
```
(ref:cprdCharacteristics-caption) Patient characteristics by drug class. Summary statistics are presented as "% (N)" unless otherwise specified in the variable name.

@@

@@

The distribution of baseline characteristics across the remaining seven drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table). Note due to the experimental design, the median year of entry is expected to be later for those not prescribed an LRA, as this exposure group is more likely to include those who entered into the cohort towards the end of study window (as these people had less follow-up time in which to be prescribed an LRA).

&nbsp;<!----------------------------------------------------------------------->  
(ref:cprdSSA-caption) Participants who stopped, switched or added treatments by initial treatment type.

@@

```{r cprdSSA-table, message=FALSE, results="asis", echo = FALSE}
```
&nbsp;<!----------------------------------------------------------------------->

@@

@@

For each outcome, the overall "Any drug" estimate was driven by the statin subgroup, based on its large size relative to the other drug classes.

@@

My results show litte evidence was found for an effect of lipid-regulating agents on probable (`r probad_text`) and possible (`r possad_text`) Alzheimer's disease when compared to no treatment, with the sole exception of fibrates on probable Alzheimer's disease (`r probad_fib_text`).

@@

In contrast to the findings for Alzheimer's disease outcomes, lipid-regulating agents were associated with an increased risk of a subsequent diagnosis of vascular dementia (`r vasdem_text`) or other dementias (`r othdem_text`). Again this effect was driven mainly by the statin subgroup, but there was some evidence that ezetimibe was associated with an increased risk of vascular (`r vasdem_eze_text`) and other (`r othdem_eze_text`) dementia.

@@

<!----------------------------------------------------------------------------->
(ref:completeCaseFig-cap) Comparison of analyses using the complete case versus imputed cohorts.
